Predicting how climate change will affect terrestrial mammals' movement and habitat selection in British Columbia, Canada

### 2025-01-17 writing/manuscript.Rmd
rewrote intro, updated methods

---
title: "Predicting how climate change will affect terrestrial mammals' movement and habitat selection in British Columbia, Canada"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Kirk Safford
    institute: bcp
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'freshwater-biology.csl'
# csl: 'nature.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

```{r, echo=FALSE}
abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move. These changes in species' movement behavior have had cascading effects on individuals' fitness and habitat selection with consequences at the population, community, and ecosystem levels. Previous research has quantified changes in mammalian movement behavior, but we are lacking an estimate of the effects of climate change on fine-scale movement and habitat selection. We address this by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on mammal's probability of movement, speed, and habitat selection. We then leverage the estimated effects to predict when, how, and where mammals will move in BC in the current century. We conclude with implications for energetics, encounter rates, habitat conservation, and human-wildlife conflicts."
```

@@

\noindent \textbf{Tables}: 1

\noindent \textbf{References}: 38 (updated on 2024-07-28) <!-- counted by using nature csl -->

@@

# Useful references to add {-}

* @hegel_interacting_2010: weather affects caribou calf recruitment and interacts with predator (wolf) abundance
* @elmore_implications_2017: "thermal environments should be foundational in the understanding of the habitat concept"
* @pease_ecological_2024: the scale at which we measure variables affects their estumated effect
* @hetem_responses_2014:
* rate of climate change is too fast for genetic adaptation to occur in mammals with longevities of decades
* example of suitable habitat prediction to 2050
* to read: https://doi.org/10.1186/s40665-016-0024-1

@@

<!-- introduce knowledge gap -->

\noindent It is well understood that mammalian energetics depend on ambient temperature [@brown_toward_2004; @taylor_energetics_1982], but we have a limited understanding of how mammals adapt their movement behavior in response to temperature [@elmore_implications_2017; but see: @milling_seasonal_2017; @montgomery_movement_2019]. Previous work has focused on energetic costs of movement [@taylor_energetics_1982; @dyer_travel_2023], range shifts [@leclerc_determinants_2021], thermoregulation [@mota-rojas_physiological_2021; @melin_effect_2023], and trait-based responses [@mccain_assessing_2019], but there remains a need for a more comprehensive understanding of how mammals will adapt their fine-scale movement and habitat selection to future changes in temperature [@intergovernmental_panel_on_climate_change_climate_2023]. Movement plasticity [@nathan_movement_2008] allows mammals to react to changes in temperature on a shorter timescale than evolutionary physiological adaptations [@levins_evolution_1974]. By changing when, how, and where they move, mammals can conserve chemical and thermal energy as well as reduce the risk of overheating [@alston_temperature_2020; @dyer_travel_2023; @attias_effects_2018; @giroux_activity_2023; @verzuh_behavioural_2023; @williams_range_2018]. We present a multi-species analysis of how mammals adapt their movement behavior in response to changes in air temperature.

<!-- movement plasticity allows for rapid responses, but how does climate change impact this? -->

Recent changes in climate have affected mammals' movement rates and habitat selection [@deb_modelling_2020], but it is still unclear how changes in the current century will affect their fine-scale movement behavior and spatial distribution by the end of the century. In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures and severe forest fires [@intergovernmental_panel_on_climate_change_climate_2023; @zurowski_summer_2023]. Over the next 100 years, these changes will continue to affect mammals' fitness, movement behavior, and habitat selection as they cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and frequency and intensity of extreme events [@bunnell_global_2011] along with anthopogenic pressure [@sih_evolution_2011; @weststrate_how_2024].

<!-- implications for ecology: encounter rates, energetic expenditure, range shifts, consequences for conservation, etc -->

Understanding animals' responses to temperature is essential for anticipating their response to climate change in the coming decades and the cascade of consequences such changes will have. In light of the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [@ref], we must understand which areas are best preserved to maximize the project's effectiveness. Predicting how climate change will affect how, when, and where animals will move is necessary for choosing which 30% of the world is most worth preserving. Changes in movement behavior impact several aspects of animals' fitness, behavior, and odds of survival, including encounter rates [with food, competitors, and predators: @martinez-garcia_how_2020], energetic expenditure [@taylor_energetics_1982], range shifts [@woo-durand_increasing_2020]. Thus, understanding the effects of temperature on animals' movement behavior and their consequences is crucial for effective conservation.

@@

This paper provides an analysis of the effects of ambient temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (*Canis lupus*, *Cervus canadensis*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over two decades of telemetry data over a large spatial range of British Columbia (Fig. \@ref(fig:bc-map)) and hourly air temperature data, we estimate how mammals altered their movement frequency, movement speed, and habitat selection in response to air temperature. We then pair the estimated responses with climate change projections to forecast changes in mammalian movement in 2100 under different climate-change scenarios [Shared Socioeconomic Pathways, see @riahi_shared_2017]. We discuss the consequences of these changes in movement behavior with regards to energetics, encounter rates, habitat conservation, and human-wildlife conflicts.

```{r bc-map, fig.cap="GPS telemetry data and 95% autocorrelated kernel density utilization distributions for the six species in this study."}
knitr::include_graphics('../figures/tels-map.png')
```

@@

```{r, eval=FALSE}
# a quick check for the table below
readRDS('../data/movement-models-speed-weights-2024-06-10.rds') %>%
  group_by(dataset_name, animal) %>%
  summarize(any_na_speed = any(is.na(speed_est)),
            all_na_speed = all(is.na(speed_est))) %>%
  group_by(dataset_name) %>%
  summarise(n_missing_any = sum(any_na_speed),
            n_missing_all = sum(all_na_speed))
```

```{r data-table}
dt_data <-
  readRDS('../models/movement-models-2024-06-06.rds') %>%
  transmute(
    Dataset = case_when(
      dataset_name == 'Canis_lupus_boreal' ~ '\\emph{Canis lupus}',
      dataset_name == 'Rangifer_tarandus_boreal' ~ '\\emph{Rangifer tarandus} (boreal)',
      dataset_name == 'Rangifer_tarandus_southern_mountain' ~ '\\emph{Rangifer tarandus} (s. mountain)',
      dataset_name == 'Puma_concolor_2' ~ '\\emph{Puma concolor}',
      dataset_name == 'Puma_concolor_4' ~ '\\emph{Puma concolor}',
      dataset_name == 'Elk in southwestern Alberta' ~ '\\emph{Cervus canadensis}',
      dataset_name == 'Oreamnos_americanus' ~ '\\emph{Oreamnos americanus}',
      dataset_name == 'Ursus_arctos_horribilis' ~ '\\emph{Ursus arctos horribilis}'),
    animal, # keep animal column for grouping later
    tel = map(tel, data.frame),
    na_speed = map_lgl(movement_model,
                       \(.m) suppressWarnings(speed(.m)$DOF == 0))) %>%
  unnest(tel) %>%
  group_by(Dataset, animal) %>%
  summarize(Start = min(as.Date(timestamp)),
            End = max(as.Date(timestamp)),
            dt = median(timestamp - lag(timestamp), na.rm = TRUE),
            na_speed = unique(na_speed),
            .groups = 'drop') %>%
  group_by(Dataset) %>%
  summarize(Start = min(Start),
            End = max(End),
            dt = as.numeric(median(dt, na.rm = TRUE)),
            n_missing = sum(na_speed),
            .groups = 'drop') %>%
  mutate(dt = round('hours' %#% dt, 2)) %>% # convert from seconds to hours
  rename('Median \U0394t (hours)' = dt,
         'NA speed' = n_missing)

knitr::kable(dt_data, format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual) and number of animals with unresolved speed (NA speed) for each of the species in this study.', escape = FALSE)
```

@@

## GPS telemetry data

\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)), and the mountain goat (*Oreamnos americanus*) locations were provided by BC Parks. All other tracking data were obtained from private collaborators. Outlier GPS locations were removed following diagnostic analyses of the distance from the median location as well as straight-line displacement, turning angle, and time interval between consecutive points. Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and high straight-line displacement, especially if antecedent and subsequent points indicated stationary behavior. The script used to clean the data and all associated custom functions are available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature)).

## Historical temperature data

\noindent Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 dataset [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)). Proximal temperature was estimated for each location by extracting the values from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [version 1.7-46, @hijmans_terra_2023] for `R`  [@r_core_team_r_2024].

@@

\noindent We modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [version 1.2.0, @fleming_ctmm_2023] for `R`. We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [416 of 433, see @fleming_fine-scale_2014; @noonan_scale-insensitive_2019]. The 18 telemetries with insufficiently fine sampling were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table)).

@@

The following two sections detail the modelling we used to estimate the effect on temperature on mammals' movement. To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year***HERE***), we fit models with and without temperature and compared the fits and model predictions of the two sets of models.

@@

\noindent We estimated the effects of temperature on mammals' movement state (moving or not) and speed (when moving) using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019 and the code chunk below] with the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017]. The first HGAM estimated the probability that an animal was moving with a binomial family of distributions and logit link function. The second HGAM estimated an animal's speed (when moving) with a gamma family of distributions and log link function. The HGAMs included random intercepts for each animal (`s(..., bs = 're')`), fixed intercepts for each species, and species-level factor smooth interactions for time of day, day of year, and temperature [`s(..., bs = 'fs')`, see model S in Figure 4 of @pedersen_hierarchical_2019]. Additionally, the models had three tensor product interaction terms (`ti()`) for each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year. These three terms accounted for: (1) seasonal changes in day length, (2) hourly changes in the response to temperature (e.g., changes in nocturnality), and (3) seasonal changes in the response to temperature (e.g., changes in coats and migration timing). The HGAMs accounted for the cyclicity of time of day and day of year using cyclic cubic splines [`bs = 'cc'`, @wood_generalized_2017]. Together, the binomial HGAM and the gamma HGAM inform us on an animal's long-term average speed, since it is the product of the probability the animal is moving and its average speed when moving. We fit the models with fast REML (`'fREML'`) and discretized covariates (`discrete = TRUE`) to optimize computational efficiency with no appreciable losses to model performance [@wood_generalized_2015; @wood_generalized_2017-1; @li_faster_2020]. Additional details are provided in Appendix A.

@@

to do:
* check `ti` terms of RSFs

>>>>

---
title: "Predicting how climate change will affect terrestrial mammals' movement and habitat selection in British Columbia, Canada"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'freshwater-biology.csl'
# csl: 'nature.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

```{r, echo=FALSE}
abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on individuals' fitness and habitat selection at the population, community, and ecosystem levels. Previous research has quantified the effects of temperature on mammalian movement behavior, but few have quantified the effects of future climate change on fine-scale movement and habitat selection. We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on mammal's probability of movement, speed, and habitat selection. We then leverage the estimated effects to predict when, how, and where mammals will move in BC throughout the 21^st^ century. We conclude with implications for energetics, encounter rates, habitat conservation, and human-wildlife conflicts. ***MISSING RESULTS AND CONCLUSIONS***"
```

@@

\noindent \textbf{Tables}: 3

\noindent \textbf{References}: 73 (updated on 2025-01-16) <!-- counted by using nature csl -->

@@

@@

<!-- movement is a convenient way to respond to change -->

\noindent For most animals, movement is a rapid, low-cost, and low-risk response to a variety of stimuli [@nathan_movement_2008], including escaping predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024], searching for resources [@charnov_optimal_1976; @mezzini_how_2024], and habitat selection [@kacelnik_ideal_1992; @winter_forecasting_2024]. However, movement efficiency depends on ambient temperature, since it impacts the energetic cost of movement [@taylor_energetics_1982; @brown_toward_2004; @jahn_variations_2022], the ease of movement [e.g., snow depth and density, see @montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023]. Additionally, temperature also affects movement indirectly by affecting other drivers of movement behavior, including physiology [@abram_behavioural_2017], fitness [@schmidt_interplay_2020; @berger_climate_2018], trophic interactions [@hegel_interacting_2010], and the availability and value of resources [e.g., water and heat, see: @elmore_implications_2017; @attias_effects_2018].

<!-- changes in temperature cause changes in movement -->

Changes in temperature affect not only whether and how fast animals move but also how they move [i.e., speed and tortuosity; their movement behavior: @nathan_movement_2008; @dyer_travel_2023] and where they move to [habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023]. Thus, temperature affects not only an animal's energetic expenditure but also many other consequences of movement, including encounter rates with resources [@mezzini_how_2024], other animals [@martinez-garcia_how_2020], and the occurrence of human-wildlife conflict [@weststrate_how_2024]. These changes are exacerbated at extreme temperatures, as animals often alter their movement behaviors substantially when temperatures are excessively cold or hot [@wilson_stenotherms_2001; @berger_climate_2018; @arechavala-lopez_common_2019; @giroux_activity_2023]. But what counts as "extreme" varies across species: thermal specialists are more sensitive to small changes above or below their ideal range [@giroux_activity_2023], while temperature generalists can adapt more easily to change [@dupont_beyond_2024]. However, this grater adaptability is tested when generalists are subject to changes that are more severe [@mccain_assessing_2019], affecting many factors [e.g., temperature, phenology, and resource availability: @polazzo_measuring_2024], or too rapid to adapt to within the span of a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021]. Thus, it remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially when populations do not react to changes in environmental conditions [@sawyer_migratory_2019] or are physiologically incapable [@sherwood_adaptability_2010; @williams_range_2018].

<!-- we do not know how temperature will change -->

This uncertainty in both future conditions and their consequences on ecosystems and populations presents a need for a comprehensive understanding of the effects of temperature on the movement of animal populations as well as the consequences this will have on landscapes, biodiversity, and their adaptability and resiliency. Recent work has documented climate-change-related changes mammals' ranges [@leclerc_determinants_2021], movement behavior [@melin_effect_2023]; thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [@mccain_assessing_2019], but it is still unclear how mammals will adapt their small-scale movement and habitat selection to future changes in temperature [@intergovernmental_panel_on_climate_change_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024]. In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures and severe forest fires [@intergovernmental_panel_on_climate_change_climate_2023; @zurowski_summer_2023]. Over the next 100 years, these changes will worsen and continue to impact mammals' fitness, movement behavior, and habitat selection as they cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and frequency and intensity of extreme events [@bunnell_global_2011] along with growing anthopogenic pressure [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024].

<!-- why is understanding the present and predicting into the future necessary -->

Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to be able to prepare and respond to change in a proactive manner [e.g., @sawyer_identifying_2009; @mccain_assessing_2019]. The Canadian province of British Columbia is currently experiencing rapid widespread warming, phenological shifts, and more frequent and intense extreme events, including forest fires [@zurowski_summer_2023]. In light of the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022], we stress that decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness. Predicting how climate change will affect how, when, and where animals will move is necessary for identifying which landscapes are most valuable now as well as in the future. Failing to understand how changes in temperature will affect mammalian movement behavior and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.

@@

This paper provides an analysis of the effects of spatiotemporally proximal temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (*Canis lupus*, *Cervus canadensis*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over two decades of telemetry data over a large spatial range of British Columbia (Fig. \@ref(fig:bc-map)) and hourly air temperature data, we estimate how mammals altered their movement frequency, movement speed, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast changes in mammalian movement in the current century under different climate-change scenarios [Shared Socioeconomic Pathways, see @riahi_shared_2017]. We discuss the consequences of these changes in movement behavior with regards to energetics, encounter rates, habitat conservation, and human-wildlife conflicts while disentangling the direct and indirect effects of temperature on the movement behavior of terrestrial mammals.

```{r bc-map, fig.cap="GPS telemetry data for the six species in this study."}
knitr::include_graphics('../figures/tels-map.png')
```

@@

```{r, eval=FALSE}
# a quick check for the table below
readRDS('../data/movement-models-speed-weights-2024-06-10.rds') %>%
  group_by(dataset_name, animal) %>%
  summarize(n = n_distinct(animal),
            any_na_speed = any(is.na(speed_est)),
            all_na_speed = all(is.na(speed_est))) %>%
  group_by(dataset_name) %>%
  summarise(n_missing_any = sum(any_na_speed),
            n_missing_all = sum(all_na_speed))
```

```{r data-table}
dt_data <-
  readRDS('../models/movement-models-2024-06-06.rds') %>%
  transmute(
    Species = case_when(
      dataset_name == 'Canis_lupus_boreal' ~ '\\emph{Canis lupus}',
      dataset_name == 'Rangifer_tarandus_boreal' ~ '\\emph{Rangifer tarandus} (boreal)',
      dataset_name == 'Rangifer_tarandus_southern_mountain' ~ '\\emph{Rangifer tarandus} (s. mountain)',
      dataset_name == 'Puma_concolor_2' ~ '\\emph{Puma concolor}',
      dataset_name == 'Puma_concolor_4' ~ '\\emph{Puma concolor}',
      dataset_name == 'Elk in southwestern Alberta' ~ '\\emph{Cervus canadensis}',
      dataset_name == 'Oreamnos_americanus' ~ '\\emph{Oreamnos americanus}',
      dataset_name == 'Ursus_arctos_horribilis' ~ '\\emph{Ursus arctos horribilis}'),
    animal, # keep animal column for grouping later
    tel = map(tel, data.frame),
    na_speed = map_lgl(movement_model,
                       \(.m) suppressWarnings(speed(.m)$DOF == 0))) %>%
  unnest(tel) %>%
  group_by(Species, animal) %>%
  summarize(Start = min(as.Date(timestamp)),
            End = max(as.Date(timestamp)),
            dt = median(timestamp - lag(timestamp), na.rm = TRUE),
            animal = unique(animal),
            na_speed = unique(na_speed),
            .groups = 'drop') %>%
  group_by(Species) %>%
  summarize(Start = min(Start),
            End = max(End),
            dt = as.numeric(median(dt, na.rm = TRUE)),
            Animals = n_distinct(animal),
            n_missing = sum(na_speed),
            .groups = 'drop') %>%
  mutate(dt = round('hours' %#% dt, 2)) %>% # convert from seconds to hours
  rename('\U0394t (hours)' = dt,
         'NA speed' = n_missing)

knitr::kable(dt_data, format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual), number of animals (Animals), and number of animals with unresolved speed (NA speed) for each of the species in this study.', escape = FALSE)
```

@@

## Datasets

This manuscript leverages four main datasets: (1) a collection of GPS telemetry data; (2) historical hourly reanalyzed air temperature from the ERA5 dataset from the European Center for Medium-range Weather Forecasting; (3) resource rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs, see @riahi_shared_2017]. We detail the data sources and analyses in the sections below, and more details are available in appendices A (estimating how temperature affects movement frequency and speed when moving), B (fitting Hierarchical Resource Selection Functions), and C (simulating hourly temperature data to 2100 using the four SSPs).

### GPS telemetry data

\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)). Clearly erroneous GPS locations were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points. Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behavior. The script used to clean the data and all associated custom functions are available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).

### Historical and projected temperature data

\noindent Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 dataset [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)). Proximal temperature was estimated for each location by extracting the value from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [version 1.7-46, @hijmans_terra_2023] for `R` [@r_core_team_r_2024].

We obtained rasters of projected monthly average temperature at a 0.08$^\circ$ resolution in BC from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`. Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for BC from 1998 to 2023 [extremes included, see @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year. We then modeled the estimated variance using a GAM for location and scale [GAMLS, see: @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017]. The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (within each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month. The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space. We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data. Additional details are available in Appendix C.

```{r, eval=FALSE, echo=FALSE}
m_gammals <- gam(list(
  # linear predictor for the mean parameter
  monthly_var ~
    s(monthly_mean, bs = 'cr', k = 10) +
    s(month, bs = 'cc', k = 10) + # using a cyclic cubic spline basis
    s(long, lat, bs = 'ds', k = 500) +
    ti(long, lat, month, bs = c('ds', 'cc'), d = c(2, 1), k = c(100, 10)),
  # linear predictor for the scale parameter
  ~ s(monthly_mean, bs = 'cr', k = 10) +
    s(month, bs = 'cc', k = 10) +
    s(long, lat, bs = 'ds', k = 100)),
  family = gammals(),
  data = d,
  method = 'REML',
  knots = list(month = c(0.5, 12.5)))
```

We simulated hourly variation in weather by assuming temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the Gamma GAMLS. We then predicted changes in movement behavior and habitat selection as a function of the simulated temperature using the HGAMs and HRSFs. For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantile by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile (Appendix C).

### Resource rasters

\noindent We estimated percent forest cover and distance from water using the rasters created by @tuanmu_global_2014. We calculated total forest cover by summing the temporally static rasters of evergreen/deciduous needleleaf trees, evergreen broadleaf trees, deciduous broadleaf trees, and mixed/other trees (raster classes 1-4, respectively). We converted the raster of percent cover of open water (class 12) to a binary raster of pixels with water (cover > 0%) or without water (cover = 0%) and then calculated each pixel's distance from the nearest pixel with water using the `distance()` function from the `terra` package. Finally, we obtaied two digital elevation models using the `get_elev_raster()` function from the `elevatr` package [version 0.99.0, @hollister_elevatr_2023]: We used a raster with a zoom of 6 (0.009$^\circ$) for model fitting and one with a zoom of 3 (0.08$^\circ$) for downloading climate change projections via `climatenaR`. All final rasters and code are available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).

@@

\noindent We modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [version 1.2.0, @fleming_ctmm_2023] for `R`. We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433, see @fleming_fine-scale_2014; @noonan_scale-insensitive_2019]. The 18 telemetries with insufficiently fine sampling were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table)).

@@

\noindent The following sections detail the statistical framework and modelling we used to estimate the effect on temperature on mammals' movement. To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the deviance explained, model predictions, and AIC values of the two sets of models.

### Disentangling direct and indirect effects

\noindent In this manuscript, we separate the effects of temperature on mammalian movement into indirect and direct effects. We call "indirect" all effects that can be intervened on by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022], and we refer to all other effects as "direct". For example, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can thus mitigate these indirect effects of temperature on movement behavior by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability. In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022], so it is not possible to mitigate the effects of temperature on movement behavior (other than changing temperature directly). We provide Directed Acyclical Graphs in Appendix A (modeling the effects of temperature on movement rates), and Appendix B (modeling the effects of temperature on habitat selection).

@@

\noindent We estimated the effects of temperature on mammals' movement state (moving or not) and speed (when moving) using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019 and the code chunk below] with the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017]. The first HGAM estimated the probability that an animal was moving with a binomial family of distributions and logit link function. The second HGAM estimated an animal's speed (when moving) with a gamma family of distributions and log link function. The HGAMs included random intercepts for each animal (`s(..., bs = 're')`), fixed-effect intercepts for each species, and species-level `by` smooths for time of day, day of year, and temperature [`s(..., by = species)`, see model I in Figure 4 of @pedersen_hierarchical_2019]. Additionally, the models had three tensor product interaction terms (`ti()`) for each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year. These three terms accounted for: (1) seasonal changes in day length, (2) hourly changes in the response to temperature (e.g., changes in nocturnality), and (3) seasonal changes in the response to temperature (e.g., changes in coats and migration timing). The HGAMs accounted for the cyclicity of time of day and day of year using cyclic cubic splines [`bs = 'cc'`, @wood_generalized_2017]. Together, the binomial HGAM and the gamma HGAM inform us on an animal's long-term average speed, since it is the product of the probability of moving and its average speed when moving. We fit the models with fast REML (`'fREML'`) and discretized covariates (`discrete = TRUE`) to optimize computational efficiency with no appreciable losses to model performance [@wood_generalized_2015; @wood_generalized_2017-1; @li_faster_2020]. Additional details are provided in Appendix A.

<!-- \clearpage -->

@@

```{r hrsfs, fig.cap="\\textbf{Temeperature strongly affects mammals' habitat selection.} Estimated resourse selection strength for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature. The grey contours indicate the extent of each speices' observed locations."}
knitr::include_graphics('../figures/hrsf-surface-plots.png')
```
